---
author: Annie Gao
title: "transit_ps6"
output: html_document
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
#install.packages("statar")
library(statar)
#install.packages("readxl")
library(readxl)
library(tibble)
library(lubridate)
#install.packages("zoo")
library(zoo)
```

```{r echo=TRUE, warning=FALSE}
func_abs_related <- read_csv("/project2/ppha30531/axgao1/func_abs_related.csv")
hrs <- read_csv("/project2/ppha30531/axgao1/hrs.csv")
ot_fund <- read_csv("/project2/ppha30531/axgao1/ot_fund.csv")
timekeep <- read_csv("/project2/ppha30531/axgao1/timekeep.csv")
weather <- readxl::read_excel("/project2/ppha30531/axgao1/Weather Events Hol.xlsx")
```

#Problem Set 5 Solutions

```{r, echo = FALSE, include = FALSE}
require(statar)
ot_fund$Payroll.Hours <- winsorize(ot_fund$Payroll.Hours, probs=c(0,0.99))
ot_fund$Overtime.Amount <- winsorize(ot_fund$Overtime.Amount, probs=c(0,0.99))
```

```{r echo=FALSE, warning=FALSE, include = FALSE}
df_list <- list(func_abs_related, hrs, ot_fund, timekeep, weather)
name <- c('func_abs_related', 'hrs', 'ot_fund', 'timekeep', 'weather')
dfs = data.frame(name)

num_rows <- unlist(lapply(df_list, nrow))
dfs$num_rows <- num_rows

start_date <- as.Date(c(NA,
                        min(as.Date(hrs$Calendar.Full.Date)),
                        min(as.Date(ot_fund$Payroll.End.Date)), 
                        min(as.Date(timekeep$Calendar.Full.Date)),
                        min(as.Date(weather$OBSERVATION_DATE))),
                      origin="1970-01-01")
dfs$start_date <- start_date


end_date <- as.Date(c(NA,
                      max(as.Date(hrs$Calendar.Full.Date)),
                      max(as.Date(ot_fund$Payroll.End.Date)),
                      max(as.Date(timekeep$Calendar.Full.Date)),
                      max(as.Date(weather$OBSERVATION_DATE))),
                    origin="1970-01-01")
dfs$end_date <- end_date
```

```{r echo=FALSE, warning=FALSE, include = FALSE}
sum_df <- function (df, name) {
  df_name <- rep(name, length(names(df)))
  variable <- names(df)
  share_missing <- unlist(lapply(df, function(x) floor(mean(is.na(x))*1000)/1000))
  mean <- unlist(lapply(df, function(x) mean(x)))
  median <- unlist(lapply(df, function(x) median(x)))
  return(data.frame(df_name, variable, share_missing, mean, median))
}

sum_far <- sum_df(func_abs_related, 'func_abs_related')
sum_hrs <- sum_df(hrs, 'hrs')
sum_otf <- sum_df(ot_fund, 'ot_fund')
sum_tk <- sum_df(timekeep, 'timekeep')
sum_w <- sum_df(weather, 'weather')

options("scipen"=100, "digits"=3)
bind_rows(sum_far, sum_hrs, sum_otf, sum_tk, sum_w) 
```

```{r echo=FALSE, warning=FALSE, include = FALSE}
require(lubridate)
hrs_tidy <- hrs %>% 
  select(-Regular...40.Total) %>%
  mutate(week = cut(Calendar.Full.Date, "week"))

hrs_tidy <- hrs_tidy %>% 
  rename(earn_cd = Payroll.Earning.Code.Description) %>%
  mutate(earning_code = ifelse(earn_cd == "BUILT IN O.T.","hrs_built_in_ot",
                               ifelse(earn_cd == "OVERTIME PAY (1.5)","hrs_ot_15",
                                      ifelse(earn_cd == "OVERTIME PAY (2.0)", "hrs_ot_20",
                                             ifelse(earn_cd == "DIFFERENTIAL PAY OT", "diff_pay", "hrs_reg_pay"))))) %>%
  select(-earn_cd) %>% spread(earning_code,Payroll.Hours) 

# to account for employees who have multiple job titles
hrs_tidy <- hrs_tidy %>% 
  group_by(week, Employee.ID) %>% 
  summarise(hrs_built_in_ot = sum(hrs_built_in_ot, na.rm = TRUE),
            hrs_ot_15 = sum(hrs_ot_15, na.rm = TRUE),
            hrs_ot_20 = sum(hrs_ot_20, na.rm = TRUE),
            diff_pay = sum(diff_pay, na.rm = TRUE),
            hrs_reg_pay = sum(hrs_reg_pay, na.rm = TRUE),
            Hourly.Rate = max(Hourly.Rate, na.rm=TRUE))
```

```{r, echo=FALSE, warning=FALSE, include = FALSE}
top3_sched <- timekeep %>% 
  filter(Absence.Type=='Scheduled') %>%
  count(TKS.Earnings.Description, sort = TRUE) %>%
  head(3)

top3_unsched <- timekeep %>% 
  filter(Absence.Type=='Unscheduled') %>%
  count(TKS.Earnings.Description, sort = TRUE) %>%
  head(3)
```

```{r echo=FALSE, warning=FALSE, include = FALSE}
tk <- timekeep %>% spread(Absence.Type,Total...of.Days, fill = 0)
timekeep_week <- tk %>%
  group_by(week = cut(Calendar.Full.Date, "week"), Employee.ID) %>%
  summarise(Unscheduled = sum(Unscheduled), Scheduled = sum(Scheduled))

timekeep_week$Total.Absences <- timekeep_week$Unscheduled + timekeep_week$Scheduled
```

```{r echo=FALSE, warning=FALSE, include = FALSE}
ht_emp_weeks <- nrow(hrs_tidy)
hrs_tidy$week <- as.Date(hrs_tidy$week)

timekeep_emp_weeks <- nrow(timekeep_week)
timekeep_week$week <- as.Date(timekeep_week$week)

week_intersect <- intersect(hrs_tidy$week, timekeep_week$week)

first_week <- min(week_intersect)
last_week <- max(week_intersect)
```

```{r echo=FALSE, warning=FALSE, include = FALSE}
hrs_tidy_week <- hrs_tidy %>% filter(between(week, first_week, last_week))
hrs_tidy_week$week <- as.Date(hrs_tidy_week$week)
timekeep_week <- timekeep_week %>% filter(between(week, first_week, last_week))
timekeep_week$week <- as.Date(timekeep_week$week)
```

```{r echo=FALSE, warning=FALSE, include = FALSE}
employee_weeks <- left_join(hrs_tidy_week, timekeep_week, by = c("Employee.ID", "week"))

employee_weeks$Total.Absences[is.na(employee_weeks$Total.Absences)] <- 0

employee_weeks$logical <- ifelse(employee_weeks$Total.Absences>0, 1, 0)

employee_weeks$hrs_built_in_ot[is.na(employee_weeks$hrs_built_in_ot)] <- 0
employee_weeks$hrs_ot_15[is.na(employee_weeks$hrs_ot_15)] <- 0
employee_weeks$hrs_ot_20[is.na(employee_weeks$hrs_ot_20)] <- 0
employee_weeks$diff_pay[is.na(employee_weeks$diff_pay)] <- 0
employee_weeks$hrs_reg_pay[is.na(employee_weeks$hrs_reg_pay)] <- 0
```



#PROBLEM SET 6 BEGINNING

#3 STA Insufficient headcount

##1
```{r}
abs_ot_funcdescr <- inner_join(ot_fund, func_abs_related, 
                               by = c("Payroll.Function.Description")) %>%
  select(Fiscal.Year, 
         Absence.Non.Absence,
         Payroll.RollUp.Department.Description, 
         Payroll.Job.Title, 
         Payroll.Function.Description, 
         Payroll.Hours, 
         Overtime.Amount, 
         Employee.ID)

abs_ot_funcdescr %>% 
  filter(Fiscal.Year == 2016 & Absence.Non.Absence == "Absence Related") %>%
  group_by(Payroll.RollUp.Department.Description) %>%
  summarise(total_dollars = sum(Overtime.Amount)) %>%
  arrange(desc(total_dollars)) %>%
  head(5)
```

The departments billing the most absence-related overtime overall in dollar terms are BusTrans, Heavy Rail, BusMaint, Rail Mntnc, and Light Rail.

##2
```{r, error = TRUE, warning = FALSE}
ot_dept_abs <- inner_join(hrs, abs_ot_funcdescr, by = c("Employee.ID", "Payroll.Job.Title"))

ot_dept_abs %>%
  group_by(Payroll.RollUp.Department.Description, Absence.Non.Absence) %>%
  summarise(ot_hrs = sum(Payroll.Hours.x, na.rm = TRUE)) %>%
  mutate(prop_abs_OT = ot_hrs/sum(ot_hrs)) %>%
  arrange(desc(prop_abs_OT)) %>%
  head(3)
```

The top 3 departments working the highest proportion of their hours as absence-related overtime are AFC1.0, Cnstr Mgmt, and Constr E&M. These departments worked all of their hours as absence-related overtime. This approach could be slightly incorrect because not all of the Payroll.Function.Description are matched in the hrs dataframe resulting in possible NAs. 

#4 STA Shenanigans

##1
```{r}
mean(employee_weeks$logical)
```

The share of employee-weeks with an absence is 28.8 percent. This estimate is extremely high since external benchmarks show absence rates to be below 5%.

##2
```{r}
#build double dip binary indicator
employee_weeks$all_ot <- employee_weeks$hrs_built_in_ot + 
  employee_weeks$hrs_ot_15 + 
  employee_weeks$hrs_ot_20 + 
  employee_weeks$diff_pay +
  employee_weeks$hrs_reg_pay

employee_weeks$dbldip <- ifelse(employee_weeks$logical == 1 & 
                                  employee_weeks$all_ot > 0, 1, 0)

#setup to filter on year
employee_weeks$year <- year(employee_weeks$week)

employee_weeks %>% 
  filter(year == 2015) %>% 
  group_by(Employee.ID) %>% 
  summarise(weeks_worked = n(), total_dbldip_emp = sum(dbldip)) %>% #weeks_worked measures how many times each employee showed by row = number of weeks worked
  filter(weeks_worked >= 26) %>% 
  mutate(share_weeks_dbldip = total_dbldip_emp/weeks_worked) %>%
  arrange(desc(share_weeks_dbldip)) %>%
  head(5)
```

The top 5 employees with the highest share of weeks double-dipping are Employee 8990, 9743, 14578, 15440, and 15747. They all have 100% double dipping rates. For all the 26 weeks or more that they worked, they doubled dipped in every week they worked.

#5 STA Snow

##1
```{r}
snow_storm <- weather %>%
  mutate(sum = PRECIPITATION + lag(PRECIPITATION)) %>%
  filter(TYPE == "S") %>%
  arrange(desc(sum)) %>%
  head(3)

snow_storm$name <- c("Storm1", "Storm2", "Storm3")
```

Using lag, the three biggest snowstorm dates as the first date are 2011-02-26, 2011-01-12, and 2013-02-09. The precipitation from these dates are sums of the previous date and the date shown for a full 48 hour span. For 2011-02-26, it seems that the day after (the 27th) had the largest accumulation.

##2
```{r}
#get all absences on each day and then filter out dates of interest and create range around them
snow_absence <- timekeep %>%
  group_by(Calendar.Full.Date) %>%
  summarise(tot_absence = sum(Total...of.Days)) %>%
  filter(between(Calendar.Full.Date, as.Date("2011-02-26") - days(5), 
                 as.Date("2011-02-26") + days(5)) |
          between(Calendar.Full.Date, as.Date("2011-01-12") - days(5), 
                  as.Date("2011-01-12") + days(5)) |
          between(Calendar.Full.Date, as.Date("2013-02-09") - days(5), 
                  as.Date("2013-02-09") + days(5)))

#first name ranges by date of storm
snow_absence <- snow_absence %>% 
  mutate(storm_dates = ifelse(between(Calendar.Full.Date, as.Date("2011-02-26") - days(5), 
                                as.Date("2011-02-26") + days(5)), "2011-02-26",
                              ifelse(between(Calendar.Full.Date, as.Date("2011-01-12") - days(5), 
                                as.Date("2011-01-12") + days(5)), "2011-01-12", "2013-02-09"))) 

#then name storms for group/color for plotting
snow_absence$names <- ifelse(snow_absence$storm_dates == "2011-02-26", "Storm1",
                             ifelse(snow_absence$storm_dates == "2011-01-12", "Storm2", 
                                    "Storm3"))

#create range for plotting x axis
snow_absence <- snow_absence %>% 
  mutate(dates_range = Calendar.Full.Date - as.Date(storm_dates))

#plot
ggplot(snow_absence, aes(x = dates_range, y = tot_absence)) +
  geom_line(aes(group = names, color = names))
```

##3
```{r}
#from original weather df, find total precipitation for each possible week
#that 3 storms coud have occurred
weather_storm_weeks <- weather %>%
  mutate(poss_week = floor_date(OBSERVATION_DATE, unit = "week")) %>%
  group_by(poss_week) %>%
  summarise(tot_precip = sum(PRECIPITATION, na.rm = TRUE)) 

#create vector of week the 3 storms occurred on
storm_weeks_vec <- c(floor_date(as.Date("2011-02-26"), "week"),
                     floor_date(as.Date("2011-01-12"), "week"),
                     floor_date(as.Date("2013-02-09"), "week"))

#create indicator for whether storm occurred in that week
weather_storm_weeks <- weather_storm_weeks %>% 
  mutate(week_w_storm = ifelse(as.Date(poss_week) %in% storm_weeks_vec, 1, 0))

#get months of all weeks and filter out Jan without precipitation
#or week storm occurred
weather_storm_weeks <- weather_storm_weeks %>%
  mutate(month = month(poss_week)) %>%
  filter((month == 1  & tot_precip == 0) | week_w_storm == 1) %>%
  mutate(storm_week = as.Date(poss_week))

#find weeks in otfund that are same weeks of storms and join two dfs
ot_fund_weeks <- ot_fund %>%
  mutate(weeks = floor_date(Payroll.End.Date, unit = "week"))
  
storm_weeks_ot <- left_join(weather_storm_weeks, ot_fund_weeks, 
                            by = c("storm_week" = "weeks")) 

storm_weeks_ot %>% 
  group_by(storm_week, week_w_storm) %>%
  summarise(tot_ot_amt = sum(Overtime.Amount))
```

Overtime costs in weeks of snowstorm are $1233006 for the week of 2013-02-03. 

##4
```{r}
#days of storm
#2011-02-26
#2011-01-12
#2013-02-09

#above
storm_weeks_vec <- c(floor_date(as.Date("2011-02-26"), "week"),
                     floor_date(as.Date("2011-01-12"), "week"),
                     floor_date(as.Date("2013-02-09"), "week"))

#convert weeks to dates in week and merge vector in employee_weeks
days_vec1 <- seq(as.Date("2011-02-20"), as.Date("2011-02-27"), by = "days")
days_vec2 <- seq(as.Date("2011-01-09"), as.Date("2011-01-16"), by = "days")
days_vec3 <- seq(as.Date("2013-02-03"), as.Date("2013-02-10"), by = "days")
days_vec <- as.vector(rbind(days_vec1, days_vec2, days_vec3))

#convert weeks to poss_weeks to match storm weeks
employee_weeks$poss_weeks <- floor_date(employee_weeks$week, "week")

#creates new df where all employee info pertain to storm weeks
employee_weeks_storm <- employee_weeks %>%
  group_by(poss_weeks) %>%
  mutate(week_indicator = ifelse(poss_weeks %in% days_vec, 1, 0)) %>%
  filter(week_indicator == 1)

#group_by
employee_weeks_storm %>%
  group_by(poss_weeks) %>%
  mutate(all_dates = days(between(poss_weeks, as.Date("2011-02-20"), as.Date("2011-02-27"))))

#multiply hrs by 1.5
employee_weeks_storm$storm_cost <- employee_weeks_storm$Hourly.Rate * 1.5

#this would make more sense if I could change weeks to dates in employee_weeks
#and then filter the dates that match storm dates and then group by storm dates
employee_weeks_storm %>%
  group_by(poss_weeks)
```
